### 2018-06-20 ident/docs/draft_4.tex
changes to intro (based on WRC comments) and results

	Constraint-based models (CBMs) of metabolism are used to improve our understanding of metabolism by representing it as a stoichiometric network of reactions \parencite{Bordbar2014a}. The ability of CBMs to shine light on the nonintuitive interactions that govern cellular metabolism is leveraged to engineer and asses the impact of designs that alter the ability of a cell to grow, or produce a desired metabolite \parencite{Maia2016}. However, in CBMs, metabolism is assumed to operate under a pseudo steady state. Consequently, the metabolite concentrations within the metabolic network are assumed to be constant, and changes in metabolite concentrations are not modeled. Furthermore, since CBMs represent metabolism using only the stoichiometry of its constituent reactions, they do not account for the various non-catalytic regulatory interactions that are also responsible for metabolic function. These shortcomings prevent CBMs from being used to fully understand the steady state as well as the dynamic characteristics of metabolic networks.

	In contrast, the effects of regulatory interactions and changes in metabolite concentrations on different characteristics of metabolism can be studied using kinetic models of metabolism \parencite{Saa2017}. These models account for changes in metabolite concentrations subject to thermodynamic and regulatory constraints that underly metabolic networks in addition to their stoichiometry \parencite{Link2014}. Kinetic models can not only help us better understand lesser known and understood characteristics of metabolism like bistability \parencite{Kotte2014}, and their role in human health, but can also improve predictions about the impact of engineering design perturbations on metabolism, and help propose alternative designs to achieve metabolite production goals \parencite{Khodayari2016}.

	Kinetic models differ from CBMs in their use of mechanistic enzyme kinetics to model the fluxes within a metabolic network \parencite{Srinivasan2015,Saa2017}. The use of kinetic models requires information on the enzyme kinetic rate laws that will be used to model the fluxes, as well as numerical values for the parameters used in these rate laws. Analyzing the ability of a metabolic network to exhibit dynamic characteristics like multiple steady states and oscillations, irrespective of the structure of the network, is one example where kinetic rate laws and parameter values play a crucial role \parencite{Srinivasan2017}.

	Despite their importance, the parameterization of kinetic models is still a problem for which solutions are a subject of debate within the modeling community. Typically, enzyme kinetic rate laws are parameterized based on in vitro observations of enzyme activity, as opposed to observations made under in vivo conditions \parencite{Heijnen2005,Smallbone2007}. However, some researchers have questioned their relevance for gleaning information on the dynamics of metabolism under in vivo conditions \parencite{Heijnen2005,Heijnen2013}. On the other hand, some reports have shown that despite the large uncertainties associated with parameters estimated based on in vivo experimental data \parencite{Link2014}, in vitro parameter estimates are a reasonable approximation of values that would be applicable under in vivo conditions \parencite{Davidi2016}.

@@

	However, the importance of model parameter identifiability, i.e., a necessary condition to estimate unique kinetic parameter values from experimental data, is often overlooked \parencite{Rodriguez-Fernandez2006,Berthoumieux2013}. Unique parameter estimates from experimental data can only be obtained for structurally and practically identifiable parameters.

	Briefly, it should be possible to obtain unique estimates for structurally identifiable parameters, just based on the model structure and parameterization, irrespective of the type and nature of experimental data available. If parameters cannot be uniquely estimated due to their nonlinear relationships with other parameters (redundant model parameterization), or due to the lack of a relationship between the measured concentrations/fluxes and the unmeasurable concentrations/fluxes (presence of unobservable system states) modeled by the parameters, then these parameters are said to be structurally non-identifiable.

	If the experimental data used for parameter estimation is informative so as to facilitate obtaining unique parameter estimates, or the quantification of uncertainties in the parameter estimates, then the parameters are said to be practically identifiable. However, if the ability to estimate unique parameter values is compromised due to the inability of the available data to capture the requisite information needed to estimate the parameters in the modeled system, or the uncertainty in parameter estimates is not quantifiable, the parameters are said to be practically non-identifiable.

	Authors have proposed ways to assess parameter identifiability by proposing approximate kinetic models of metabolism that utilize empirical enzyme kinetic rate laws with parameters that have physical significance, and are identifiable \parencite{Heijnen2005,Smallbone2007}. Significant work has also been done towards the development of methods for structural identification of parameters in kinetic models of metabolism \parencite{Nikerel2009,Berthoumieux2013,Raue2014}. \textcite{Chis2011b} provide a comparison of the computational performance and implementation aspects of some of these methods.

	Methods to improve practical identifiability through a priori experimental design have also been developed, with a focus on kinetic models of metabolism \parencite{Gadkar2005a,Rodriguez-Fernandez2006,Vanlier2014a,Raue2014}. Some of these methods are limited by their applicability to approximate kinetic models only \parencite{Nikerel2009,Berthoumieux2013}, while some others suffer from computational limitations when applied to kinetic models of large metabolic networks \parencite{Gadkar2005a,Raue2014}. 	
	Assessing the identifiability (structural or practical) of parameters using any of the aforementioned or other methods is done under the assumption that only dynamic time course concentration data is available and can be used for parameter estimation. The ability to use steady state experimental data to assess parameter identifiability is rarely addressed, and very few methodology have been proposed to fill this gap \textcolor{red}{(paper that proposes both steady state and dynamic data use for a priori experimental design resulting in reduced model sizes)}

	In this paper, we show that the identifiability of most parameters in a metabolic network can be done for each reaction individually using steady state fluxomics, metabolomics and proteomics data, and propose a methodology to do so. We also demonstrate how this method can facilitate the design of experiments for generating the necessary steady state data to estimate unique parameter values for all reaction fluxes in a metabolic network. In doing so, we are able to propose the number and types of perturbations that will provide the most useful data for parameter estimation. We also comment on the identifiability of different enzyme kinetic rate laws that are typically used to model fluxes in metabolic networks using steady state data. We illustrate our methodology using a small metabolic network model of glucoeneogenesis in \textit{Escherichia coli} \parencite{Kotte2014, Srinivasan2017} under the assumption that all intracellular metabolite concentrations and fluxes can be measured.

	\section{Methods}\label{sec:methods}
	In Section \ref{sec:ident_def} we present typical structures of kinetic models of metabolism and pertinent formal mathematical definitions for structural and practical identifiability. The computer-algebra system based method to assess identifiability that we have developed is described in Section \ref{sec:ident} that follows. In Section \ref{sec:degree_of_identifiability}, we define a quantitative metric to describe the identifiability of parameters, followed by a description of how the method can be used for experimental design for parameter estimation is given in Section \ref{sec:experimental_design}. We provide a description of a numerical equivalent for the computer algebra-based method to estimate parameters numerically in Section \ref{sec:numerical_method} (\textcolor{red}{should this be moved to SI?}). Finally, in Section \ref{sec:small-model} a complete description of the small metabolic network used to demonstrate the methodology that we have developed is provided.

@@

	\subsection{Numerical parameter identification and estimation}\label{sec:numerical_method}
	The identification methodology described in the previous section can be extended to be used in conjunction with an optimization-based procedure for numerical estimation of parameters from experimental data. The $\mathbb{R}^p$ nonlinear algebraic equations formulated for each flux $v_i$ (Equation \ref{eq:nonlineq}) can be written as shown in Equation (\ref{eq:numerical_opt}) below to be solved as an optimization problem. Minimization of the sum of the absolute error values is used as the objective, and the $\mathbb{R}^p$ nonlinear algebraic equations are provided as inequality constraints.

	\begin{subequations}\label{eq:numerical_opt}
		\begin{align}
		\underset{\theta, \varepsilon}{min} & \sum_{j=1}^{n_p} \left|\varepsilon_j\right|\\
		& v_{i, j} - f_j(\mathbf{x}_j,\mathbf{\theta}, \mathbf{u}_j) \le \varepsilon_j && j=\{1, 2, ..., n_p\}\\
		& \theta_l \le \theta_j \le \theta_u && j=\{1, 2, ..., n_p\}
		\end{align}
	\end{subequations}

	Notice that the parameter values are bounded above and below by $\theta_u$ and $\theta_l$, respectively. These bounds can be specified to be arbitrarily large to include all possible parameter values.

@@

	First, we find that the spread in the estimated Convenience kinetics parameter values, when $v_3$ is only locally structurally identifiable, is much larger than when there is mismatch between the model generating the data and the model that is being identified (Supplementary Figure). A more important observation is that even when the parameters are structurally identifiable in $v_3$ (achieved by assuming either $K_3^{fdp}$ or $K_3^{pep}$ as a known constant), they can at most only be locally practically identifiable. This is shown by the spread in the estimate values of the structurally identifiable parameters when steady state data based on the MWC model is used in Supplementary Figure.

@@

	Apart from the mismatch between the in vivo and the in vitro enzyme kinetic rate laws, uncertainty in experimental data also arises due to the presence of noise in the measured experimental data. This noise could be attributed to the measurement error commonly encountered in process analytics. In order to test the robustness of our methodology to practically identify parameters using steady state data with measurement errors, we used in silico experimental data with 5\% additive noise for practical identification, instead of the noise-free data that we have used so far.

	We found that in every case where parameters are structurally identifiable, the noise did not have any effect on the identifiability of the parameters or their estimated values. We also found that inclusion of all necessary data (e.g., the presence or lack thereof of enzyme concentration \textit{E} for $v_1$) can alleviate issues related to using experimental data with errors for identification and estimation: the degrees of identifiability of $V_1^{max}$ and $K_1^{ac}(ne)$ had non-zero standard deviations associated with them, but the degrees of identifiability of $k_1^{cat}$ and $K_1^{ac}$ did not. Using a similar reasoning to the earlier scenario in the presence of mismatches between in vitro and in vivo model, we can say that there is no room for any uncertainties to propagate from the noisy data when all necessary steady state information for identification is available. Thus, both $v_2$ and $v_5$ also did not show any differences in either their degree of identifiability or their estimated parameter values.

	However, for $v_3$, whose parameters are only locally structurally identifiable, we found small non-zero standard deviations in the degrees of identifiabilities (Supplementary Figure) when noisy data is used. Although this was seen due to the differences in the number of data combinations that can estimate positive values for each of the three parameters between different noisy experimental data sets, we observe that the standard deviation in the estimated parameter values for each data, between different samples of noisy experimental data, is small (Supplementary Figure).

>>>>

	Constraint-based models (CBMs) of metabolism are used to improve our understanding of metabolism by representing it as a stoichiometric network of reactions \parencite{Bordbar2014a}. The ability of CBMs to shine light on the nonintuitive interactions that govern cellular metabolism is leveraged to engineer and asses the impact of designs that alter the ability of a cell to grow, or produce a desired metabolite \parencite{Maia2016}. However, in CBMs, metabolism is assumed to operate under a pseudo steady state. Consequently, the metabolite concentrations within the metabolic network are assumed to be constant, and changes in metabolite concentrations are not modeled. Furthermore, since CBMs represent metabolism using only the stoichiometry of its constituent reactions, they do not account for the various dynamic interactions that are also responsible for metabolic function. These shortcomings prevent CBMs from being used to fully understand the steady state as well as the dynamic characteristics of metabolic networks.

	In contrast, the effects of regulatory interactions and changes in metabolite concentrations on different characteristics of metabolism can be studied using kinetic models of metabolism \parencite{Saa2017}. These models account for changes in metabolite concentrations subject to thermodynamic and regulatory constraints that underly metabolic networks in addition to their stoichiometry \parencite{Link2014}. As a result kinetic models can help us better understand lesser known and understood characteristics of metabolism like bistability \parencite{Kotte2014}, and their role in human health. They can also improve predictions about the impact of engineering design perturbations on metabolism, and help propose alternative designs to achieve metabolite production goals \parencite{Khodayari2016}.

	Kinetic models differ from CBMs in their use of mechanistic enzyme kinetics to model the fluxes within a metabolic network \parencite{Srinivasan2015,Saa2017}. The use of kinetic models requires information on the enzyme kinetic rate laws that are used to model the fluxes, as well as numerical values for the parameters used in these rate laws. In addition to the structure of the network represented by kinetic models, the parameter values play a crucial role in analyzing the ability of a metabolic network to exhibit different dynamic characteristics like multiple steady states and oscillations \parencite{Srinivasan2017}.

	Despite their importance, the parameterization of kinetic models is still a problem for which solutions are a subject of debate within the modeling community. Typically, enzyme kinetic rate laws are parameterized based on in vitro observations of enzyme activity, as opposed to observations made under in vivo conditions \parencite{Heijnen2005,Smallbone2007}. Although some researchers have questioned their relevance for gleaning information on the dynamics of metabolism under in vivo conditions \parencite{Heijnen2005,Heijnen2013}, some others have shown that in vitro parameter estimates are a reasonable approximation of values that would be applicable under in vivo conditions \parencite{Davidi2016}.

	%On the other hand, some reports have shown that despite the large uncertainties associated with parameters estimated based on in vivo experimental data \parencite{Link2014},

@@

	The importance of model parameter identifiability, i.e., a necessary condition to estimate unique kinetic parameter values from experimental data, is often overlooked \parencite{Rodriguez-Fernandez2006,Berthoumieux2013}. Unique parameter estimates from experimental data can only be obtained for structurally and practically identifiable parameters.

	To be structurally identifiable, a parameter should be capable of being uniquely estimated, irrespective of the type and nature of available experimental data. This aspect of a parameter is related to the structure of the model and its parameterization. For instance, a parameter is structurally non-identifiable if it is correlated to another model parameter due to redundant model parameterization, and consequently cannot be estimated uniquely from any available experimental data.

	To be practically identifiable, a parameter should be structurally identifiable, and it should be possible to obtain unique parameter estimates from available experimental data. Unique parameters estimates for structurally identifiable parameters cannot be determined from experimental data that contribute to practical non-identifiability of the parameter. This is attributed to the inability of the data to capture the requisite information needed for estimating the parameter.

	%Briefly, it should be possible to obtain unique estimates for structurally identifiable parameters, just based on the model structure and parameterization, irrespective of the type and nature of experimental data available. If parameters cannot be uniquely estimated due to their nonlinear relationships with other parameters (redundant model parameterization), or due to the lack of a relationship between the measured concentrations/fluxes and the unmeasurable concentrations/fluxes (presence of unobservable system states) modeled by the parameters, then these parameters are said to be structurally non-identifiable.

	%If the experimental data used for parameter estimation is informative so as to facilitate obtaining unique parameter estimates, or the quantification of uncertainties in the parameter estimates, then the parameters are said to be practically identifiable. However, if the ability to estimate unique parameter values is compromised due to the inability of the available data to capture the requisite information needed to estimate the parameters in the modeled system, or the uncertainty in parameter estimates is not quantifiable, the parameters are said to be practically non-identifiable.

	Authors have developed approximate kinetic models of metabolism to assess parameter identifiability. These models utilize empirical enzyme kinetic rate laws with parameters that have physical significance, and are identifiable \parencite{Heijnen2005,Smallbone2007}. Significant work has also been done towards the development of methods for structural identification of parameters in kinetic models of metabolism \parencite{Nikerel2009,Berthoumieux2013,Raue2014}. \textcite{Chis2011b} provide a comparison of the computational performance and implementation aspects of some of these methods. 	
	Methods to improve practical identifiability through a priori experimental design have also been developed, with a focus on kinetic models of metabolism \parencite{Gadkar2005a,Rodriguez-Fernandez2006,Vanlier2014a,Raue2014}.

	Despite the availability of these methods, the question that has not been answered is the quality and the quantity of experimental data that is required to uniquely estimate parameters in kinetic models of metabolism. In this paper, we seek to answer this question in the context of identifiability. We present a methodology to determine the conditions under which we can identify parameters, and estimate the number of experimental data sets required for the complete estimation of all parameters in a kinetic model. In doing so, we show that the number of experimental data sets required to identify all parameters is bounded by the maximum number of independent parameters in a rate law, if all concentrations and fluxes can be measured.

	We describe the framework for assessing the identifiability of parameters in kinetic models of metabolism in the Methods section. We demonstrate the utility of this methodology by illustrating its application to identify parameters in the kinetic model of a small metabolic network. In doing so, we also illustrate how some of the metrics that we have developed for assessing practical identifiability of parameters can be used to determine the number and nature of experiments required to generate informative data sets required for parameter estimation.  
	%Since some of the aforementioned methods \parencite{Gadkar2005a,Raue2014} suffer from computational limitations when applied to kinetic models of large metabolic networks, we also improve computational tractability by enabling structural and practical identification of parameters for each reaction in the network individually using steady state data only.

	%Also, the ability to use steady state experimental data to assess parameter identifiability is rarely addressed \textcolor{red}{(paper that proposes both steady state and dynamic data use for a priori experimental design resulting in reduced model sizes)}, and the assumption that only dynamic time course concentration data is available, and can be used for identifiability analysis and parameter estimation is prevalent is all of the above methods.

	%. Also, some of the aforementioned methods are limited by their applicability to approximate kinetic models only \parencite{Nikerel2009,Berthoumieux2013}, and some others

	%We seek to address these issues by first developing a framework to determine the structural and practical identifiability of parameters in kinetic models of metabolism by using only steady state concentration and flux measurements. We describe this framework in the Methods section. We also demonstrate how identifiability analysis of kinetic models of metabolism with steady state data can be computationally tractable by demonstrating how the proposed methodology can be applied to identify parameters .

	%The application of the methodology is illustrated using one flux from a small metabolic network as an example in Section \ref{sec:example}. In order to enable us to determine the

	%, our proposed methodology can be applied to a large class of kinetic rate laws that are used in the construction of kinetic models of metabolism. Some of these methods also

	%We propose to avoid these computational limitations associated with scale-up

	%We also demonstrate how this method can facilitate the design of experiments for generating the necessary steady state data to estimate unique parameter values for all reaction fluxes in a metabolic network.

	%In this paper, we show that the identifiability of , and propose a methodology to do so.  In doing so, we are able to propose the number and types of perturbations that will provide the most useful data for parameter estimation. We also comment on the identifiability of different enzyme kinetic rate laws that are typically used to model fluxes in metabolic networks using steady state data. We illustrate our methodology using a small metabolic network model of glucoeneogenesis in \textit{Escherichia coli} \parencite{Kotte2014, Srinivasan2017} under the assumption that all intracellular metabolite concentrations and fluxes can be measured.

	\section{Methods}\label{sec:methods}
	In Section \ref{sec:ident_def} we present typical structures of kinetic models of metabolism and pertinent formal mathematical definitions for structural and practical identifiability. The computer-algebra system based method to assess identifiability that we have developed is described in Section \ref{sec:ident} that follows. In Section \ref{sec:degree_of_identifiability}, we define a quantitative metric to describe the identifiability of parameters, followed by a description of how the method can be used for experimental design for parameter estimation is given in Section \ref{sec:experimental_design}. Finally, in Section \ref{sec:small-model} a complete description of the small metabolic network used to demonstrate the methodology that we have developed is provided. %We provide a description of a numerical equivalent for the computer algebra-based method to estimate parameters numerically in Section \ref{sec:numerical_method} (\textcolor{red}{should this be moved to SI?}).

@@

	%\subsection{Numerical parameter identification and estimation}\label{sec:numerical_method}
	%The identification methodology described in the previous section can be extended to be used in conjunction with an optimization-based procedure for numerical estimation of parameters from experimental data. The $\mathbb{R}^p$ nonlinear algebraic equations formulated for each flux $v_i$ (Equation \ref{eq:nonlineq}) can be written as shown in Equation (\ref{eq:numerical_opt}) below to be solved as an optimization problem. Minimization of the sum of the absolute error values is used as the objective, and the $\mathbb{R}^p$ nonlinear algebraic equations are provided as inequality constraints.

	%\begin{subequations}\label{eq:numerical_opt}
	%	\begin{align}
	%	\underset{\theta, \varepsilon}{min} & \sum_{j=1}^{n_p} \left|\varepsilon_j\right|\\
	%	& v_{i, j} - f_j(\mathbf{x}_j,\mathbf{\theta}, \mathbf{u}_j) \le \varepsilon_j && j=\{1, 2, ..., n_p\}\\
	%	& \theta_l \le \theta_j \le \theta_u && j=\{1, 2, ..., n_p\}
	%	\end{align}
	%\end{subequations}

	%Notice that the parameter values are bounded above and below by $\theta_u$ and $\theta_l$, respectively. These bounds can be specified to be arbitrarily large to include all possible parameter values.

@@

	First, we find that the spread in the estimated Convenience kinetics parameter values, when $v_3$ is only locally structurally identifiable, is much larger than when there is mismatch between the model generating the data and the model that is being identified (Supplementary Figure \ref{fig:v3_mwc_ck_values}). A more important observation is that even when the parameters are structurally identifiable in $v_3$ (achieved by assuming either $K_3^{fdp}$ or $K_3^{pep}$ as a known constant), they can at most only be locally practically identifiable. This is shown by the spread in the estimate values of the structurally identifiable parameters when steady state data based on the MWC model is used in Supplementary Figure.

@@

	%Apart from the mismatch between the in vivo and the in vitro enzyme kinetic rate laws, uncertainty in experimental data also arises due to the presence of noise in the measured experimental data. This noise could be attributed to the measurement error commonly encountered in process analytics. In order to test the robustness of our methodology to practically identify parameters using steady state data with measurement errors, we used in silico experimental data with 5\% additive noise for practical identification, instead of the noise-free data that we have used so far.

	%We found that in every case where parameters are structurally identifiable, the noise did not have any effect on the identifiability of the parameters or their estimated values. We also found that inclusion of all necessary data (e.g., the presence or lack thereof of enzyme concentration \textit{E} for $v_1$) can alleviate issues related to using experimental data with errors for identification and estimation: the degrees of identifiability of $V_1^{max}$ and $K_1^{ac}(ne)$ had non-zero standard deviations associated with them, but the degrees of identifiability of $k_1^{cat}$ and $K_1^{ac}$ did not. Using a similar reasoning to the earlier scenario in the presence of mismatches between in vitro and in vivo model, we can say that there is no room for any uncertainties to propagate from the noisy data when all necessary steady state information for identification is available. Thus, both $v_2$ and $v_5$ also did not show any differences in either their degree of identifiability or their estimated parameter values.

	%However, for $v_3$, whose parameters are only locally structurally identifiable, we found small non-zero standard deviations in the degrees of identifiabilities (Supplementary Figure) when noisy data is used. Although this was seen due to the differences in the number of data combinations that can estimate positive values for each of the three parameters between different noisy experimental data sets, we observe that the standard deviation in the estimated parameter values for each data, between different samples of noisy experimental data, is small (Supplementary Figure).
